Distances

### 2020-05-05 03_distances.Rmd
Calculate shortest path distances

---
title: "03_distances"
---

@@

For this

>>>>

---
title: "Distances"
---

@@

To calculate network-based distances between block group / tract centroids and
park boundaries, we employ the `osmnx` library for Python. This is a relatively
new package that allows users to retrieve networks from OpenStreetMap and 
compute shortest paths using Python's other network libraries. We retrieve
a walk network for New York City, and re-project it to UTM zone 18N

```py
graph = ox.graph_from_place("New York City", network_type='walk')
# project to UTM zone 18 N and simplify
graph_proj = ox.project_graph(graph)
graph_proj = nx.DiGraph(graph_proj)
```

One limitation of the network calculation is that it only works for nodes on the
network, and we can only find nearest nodes to points. Thus we need to convert
the park polygons data into points. But we can't simply have an endless number
of points because the shortest path calculations are computationally extensive.

So to make this work, we first simplify the parks polygons, convert the boundaries
to linestrings, and then sample points along the linestring.

```{r park_points, eval = FALSE}
open_spaces <- open_spaces %>%
  filter(id == "M010") %>%
  select(id)  %>%
  # some polygons are way too detailed, so we want to simplify to 100 foot resolution
  st_simplify(dTolerance = 100, preserveTopology = TRUE)
# some parks have multiple polygons. We need these to be simpler individual features 
poly_from_multi <- open_spaces %>%
  filter(st_geometry_type(geometry) == "MULTIPOLYGON") %>%
  st_cast("POLYGON")
linestrings <- open_spaces %>%
  # join the original polygons back with the simplified multipolygons
  filter(st_geometry_type(geometry) != "MULTIPOLYGON") %>%
  rbind(poly_from_multi) %>%
  # cast the polygons to a linestring of the park perimeter
  st_cast("LINESTRING", group_or_split = TRUE) 
point_samples <- linestrings %>%
  # sample points along the line, one point per 1000 feet.
  st_line_sample(density = 1/500) 
# append open space id
open_space_points <- tibble(id = linestrings$id, geometry = point_samples) %>%
  st_as_sf() %>%
  st_cast(to = "POINT")
```

```{r open_space_points, echo = FALSE}
park_points <- read_csv("data/park_points.csv") %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 32618)
```

We can illustrate what this looks like. Every park has at least one sampled point,
and some have several. Unfortunately there's not a concave polygon function to help
us remove all the internal points from the parks.

```{r simplified-point-map}
leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(data = open_spaces %>% st_transform(4326), color = "green") %>%
  addCircleMarkers(data = park_points %>% st_transform(4326), radius = 0.001,
                   color = "red") 
```

With `r nrow(open_space_points)` and `r nrow(bg_centroids)`, that suggests we
need to compute `r nrow(open_space_points) * nrow(bg_centroids)` shortest
paths. Even on a 20+ core process, this would take several weeks.  As a result,
we simplify the problem by first identifying the "closest" point of the park
by Euclidean distance and then computing the network path to that point.
This introduces some distortion where if the closest Euclidean point is across 
a river or interstate, the method calculates the network distance to *that*
point instead of the first point the person would encounter following the
network. This may end up overstating the true experienced distance.

```py
# loop through centroids for block group or tract
for bg in df.itertuples():

    # loop through parks
    for park in park_ids:
        these_points = park_points.loc[[park]]
        min_euc = float("inf")

        # loop through points associated with park
        for point in these_points.itertuples():
            # find closest park point by Euclidean distance
            dx = point.LONGITUDE - bg.LONGITUDE
            dy = point.LATITUDE - bg.LATITUDE
            euc_dist = math.sqrt(dx**2 + dy**2)

            if euc_dist < min_euc:
                min_euc = euc_dist
                closest_point = point.node

        # compute shortest path between closest_point and centroid
        try:
            length = nx.shortest_path_length(graph, source=bg.node, target=closest_point, weight='length')
        except:
            # can fail if no path exists
            length = float("inf")
```

Calculated this way, the path is calculated in meters.
